\section{Nonlinear Inequalities}

\subsection{Review problems}

\begin{enumerate}
\item Express the solutions to each of the following in interval notation.
\begin{enumerate}
\item $x^2 - 9 > 0$
\item $x^2 + 6x - 27\leq 0$
\item $2x^2 + 9x - 3\geq 2$
\item $2x^2 + 5 < 6x$
\end{enumerate}
\item Express the solutions to each of the following in interval notation.
\begin{enumerate}
\item $\sqrt{x} < \sqrt[4]{25}$
\item $(x - 1)(x - 2)(x - 3) > 0$
\item $\frac{x^2 + x + 1}{2x^2 + x - 15}\leq 0$
\item $(x + 1)(x + 2)(x + 3)(x + 4)\geq (x - 1)(x - 2)(x - 3)(x - 4)$
\end{enumerate}
\item Find the area of the region in the cartesian plane consisting of all points $(x,y)$ satisfying
\begin{equation*}
(x - 2)^2 + (y - 3)^2\geq 1\quad\text{and}\quad x^2 + y^2 - 4x - 6y\leq 3.
\end{equation*}
\item Two positive real numbers have a sum of $22$. What is their maximum possible product?
\item Carl has 20 feet of fencing to use to build a rectangular pen for his sheep. One side of the pen will be along his barn and thus does not require fencing, but the other three sides need to be entirely fenced off. What is the maximum possible area that Carl can fence off?
\item Find the minimum possible value of the expression
\begin{equation*}
4x^2 + 3y^2 - 32x + 30y - 8
\end{equation*}
when $x$ and $y$ are real numbers.
\end{enumerate}


\subsection{Challenge problems}

\begin{enumerate}[resume]
\item For any finite list of positive real numbers, the \emph{harmonic mean} is computed by taking the reciprocal of the arithmetic mean of the reciprocals, i.e. for two numbers,
\begin{equation*}
HM(x,y) = \left(\frac{x^{-1} + y^{-1}}{2}\right)^{-1}.
\end{equation*}
\begin{enumerate}
\item Compute the harmonic mean of $40$ and $60$.
\item Eggsy drives to work at a speed of $60$ miles per hour, then drives back along the same route at a speed of $40$ miles per hour. What was Eggsy's average speed for the entire round trip?
\end{enumerate}
\item In class, we stated the AM-GM inequality for two positive real numbers,
\begin{equation*}
\frac{x + y}{2}\geq\sqrt{xy}.
\end{equation*}
Throughout this problem, assume all variables represent positive real numbers.
\begin{enumerate}
\item By manipulating $\left(\sqrt{x} - \sqrt{y}\right)^2\geq 0$, prove the two-variable AM-GM inequality.
\item Use the two-variable AM-GM inequality to prove the four-variable AM-GM inequality
\begin{equation*}
\frac{w + x + y + z}{4}\geq\sqrt[4]{wxyz}.
\end{equation*}
\item Use the four-variable AM-GM inequality to prove the three-variable AM-GM inequality
\begin{equation*}
\frac{x + y + z}{3}\geq\sqrt[3]{xyz}.
\end{equation*}
\end{enumerate}
\item A rectangular box with an open top is made with a surface area of $18$ square feet. What is the maximum possible volume within the box?
\item Are there any positive real numbers $a,b,c$ satisfying the system of equations
\begin{equation*}
a^2 + ab + b^2 = 9,\quad a^2 + ac + c^2 = 16,\quad b^2 + bc + c^2 = 64?
\end{equation*}
\end{enumerate}


\newpage
\subsection{Answers}

\begin{enumerate}
\item \begin{enumerate}
\item $(-\infty, -3)\cup (3, +\infty)$
\item $[-9,3]$
\item $(-\infty, -5]\cup [1/2, +\infty)$
\item no solutions ($\emptyset$)
\end{enumerate}
\item \begin{enumerate}
\item $[0, 5)$
\item $(1,2)\cup (3, +\infty)$
\item The discriminant of $x^2 + x + 1$ is $1^2 - 4\cdot 1\cdot 1 = -3 < 0$, so $x^2 + x + 1$ is always positive or always negative. At $x = 0$, the numerator evaluates to $1$, so it must always be positive. Therefore, we are looking for when the denominator is negative,
\begin{equation*}
2x^2 + x - 15 < 0.
\end{equation*}
Factoring, we have $(2x - 5)(x + 3) < 0$, and the solution set is $\boxed{(-3, 5/2)}$.
\item Expanding both sides,
\begin{align*} 
x^4 + 10x^3 + 35x^2 + 40x + 24 &\geq x^4 - 10x^3 + 35x^2 - 40x + 24, \\
20x^3 + 80x &\geq 0, \\
x^3 + 4x &\geq 0, \\
x(x^2 + 4) &\geq 0.
\end{align*}
Since $x^2 + 4 > 0$, this is satisfied precisely when $x\geq 0$, so the solution set is $\boxed{[0, +\infty)}$.
\end{enumerate}
\item Completing the square in the second inequality yields
\begin{equation*}
(x - 2)^2 + (y - 3)^2\leq 16.
\end{equation*}
Therefore, taking the two inequalities together, the region satisfying both is the region between two concentric circles centered at $(2,3)$, one with radius $1$ and one with radius $4$. The area of this region (an \href{https://en.wikipedia.org/wiki/Annulus_(mathematics)}{annulus}) is $\pi\cdot 4^2 - \pi\cdot 1^2 = \boxed{15\pi}$.
\item If the two numbers are $x$ and $y$, then by AM-GM,
\begin{equation*}
\sqrt{xy}\leq\frac{x + y}{2} = 11.
\end{equation*}
Hence $xy\leq\boxed{121}$, and equality is attained when $x = y = 11$.
\item Let $x$ be the length of the pen parallel to the barn and let $y$ be the width perpendicular to the barn. From the given conditions, $x + 2y = 20$. Then, by AM-GM,
\begin{equation*}
\sqrt{x\cdot 2y}\leq\frac{x + 2y}{2} = 10.
\end{equation*}
This means $2xy\leq 100$, so $xy\leq\boxed{50}$. Equality is attained when $x = 2y$, in which case we have $x = 10$ and $y = 5$.
\item Completing the square,
\begin{align*} 
4x^2 + 3y^2 - 32x + 30y - 8 &= 4[x^2 - 8x + {\color{red}\boxed{\phantom{16}}}] + 3[y^2 + 10y + {\color{blue}\boxed{\phantom{25}}}] - 4{\color{red}\boxed{\phantom{16}}} - 3{\color{blue}\boxed{\phantom{25}}} - 8 \\
&= 4(x - 4)^2 + 3(y + 5)^2 - 4\cdot {\color{red}16} - 3\cdot {\color{blue}25} - 8 \\
&= 4(x - 4)^2 + 3(y + 5)^2 - 147.
\end{align*}
Since the perfect squares are always non-negative, the minimum possible value of the given expression is $\boxed{-147}$, attained when $x = 4$ and $y = -5$.
\item \begin{enumerate} 
\item $48$
\item $48$ mph
\end{enumerate}
\item \begin{enumerate}
\item Expanding the left hand side,
\begin{equation*}
x - 2\sqrt{xy} + y\geq 0,
\end{equation*}
so $x + y\geq 2\sqrt{xy}$. Dividing by $2$ on both sides yields AM-GM for two variables.
\item Let $a = (w + x)/2$ and $b = (y + z)/2$. Then using AM-GM for two variables,
\begin{align*}
\frac{w + x + y + z}{4} &= \frac{a + b}{2} \\
&\geq\sqrt{ab} \\
&=\sqrt{\frac{w + x}{2}\cdot\frac{y + z}{2}} \\
&\geq\sqrt{\sqrt{wx}\sqrt{yz}} \\
&= \sqrt[4]{wxyz}.
\end{align*}
\item Using AM-GM for four variables,
\begin{align*} 
\frac{x + y + z}{3} &= \frac{x + y + z + \frac{x + y + z}{3}}{4} \\
&\geq\sqrt[4]{xyz\cdot\frac{x + y + z}{3}} \\
&= (xyz)^{1/4}\left(\frac{x + y + z}{3}\right)^{1/4}.
\end{align*}
This gives us
\begin{equation*}
\left(\frac{x + y + z}{3}\right)^{3/4}\geq (xyz)^{1/4},
\end{equation*}
and raising both sides to the $4/3$ power proves three-variable AM-GM.
\end{enumerate}
\item Let $\ell$ be the length, $w$ be the width, and $h$ be the height of this box, and suppose the open face is one of the $\ell\times w$ faces. Then 
\begin{equation*}
\ell w + 2\ell h + 2wh = 18.
\end{equation*}
By three-variable AM-GM,
\begin{equation*}
\sqrt[3]{(\ell w)(2\ell h)(2wh)}\leq\frac{\ell w + 2\ell h + 2wh}{3} = 6.
\end{equation*}
Cubing both sides gives us $4\ell^2 w^2h^2\leq 216$, so $\ell^2 w^2h^2\leq 54$ and $\ell wh\leq\sqrt{54} = \boxed{3\sqrt{6}}$.
\item From the third equation,
\begin{equation*}
b + c = \sqrt{b^2 + 2bc + c^2}\geq\sqrt{b^2 + bc + c^2} = 8.
\end{equation*}
However, from the first and second equations,
\begin{equation*}
b\leq\sqrt{a^2 + ab + b^2} = 3\quad\text{and}\quad c\leq\sqrt{a^2 + ac + c^2} = 4.
\end{equation*}
We cannot simultaneously have $b\leq 3$, $c\leq 4$, and $b + c\geq 8$, so there are no solutions in positive real numbers to the given system of equations.
\end{enumerate}