\documentclass{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx} % Required for inserting images
\usepackage{multicol}
\usepackage[letterpaper]{geometry}
\usepackage[colorlinks]{hyperref}

\title{Algebra 1 Practice Problems}
\author{Alan Zhou}
\date{2023}



\begin{document}

\maketitle

\section{Exponents and Radicals}

\begin{enumerate}
\item (Exponent rules)
\begin{enumerate}
\item If $3^4\cdot 3^5\cdot 3^7 = 3^n$, what is $n$?
\item Compute $\dfrac{2^{16}\cdot 2^{20}}{2^{11}\cdot 2^{23}}$.
\item Compute $58492^0\cdot 6^3 + 2^8$.
\item Is $3^{-4}$ greater than or less than $0$?
\item Is $(-1/2)^{-2}$ greater than or less than $1$?
\item Compute $(-8)^{2/6}$, or if it is undefined, state so.
\item Simplify $\left(2^{\sqrt{2}}\right)^{\sqrt{2}}$.
\item Simplify $(4a^4b^{-2}c^{-6})^{-3/2}$. Your final answer should have no negative exponents.
\end{enumerate}
\item Put each of the following into simplest radical form.
\begin{enumerate}
\item $\sqrt{100}$
\item $\sqrt{8}$
\item $\sqrt{216}$
\item $\sqrt{2023}$
\item $\sqrt[3]{144}$
\item $\sqrt[4]{729}$
\item $\sqrt{50} + \sqrt{162}$
\item $\sqrt{\dfrac{2024}{2023}}$ (Your answer should not have any roots in the denominator.)
\end{enumerate}
\item Solve for $x$ in the equation
\begin{equation*}
3^{8 - 2x}\cdot 27^{x - 1} = 9^{2x + 1}.
\end{equation*}
\end{enumerate}

\newpage
\section{Expanding and Factoring}

\begin{enumerate}
\setcounter{enumi}{3}
\item Expand each of the following with the distributive property.
\begin{enumerate}
\item $xy(6x + 7y)$
\item $5a^2b(3ab - ab^2)$
\item $-3x(2 - 4x - 6x^2)$
\item $\dfrac{-8a^2b - 6ab^2 + 4a^2b^2}{2ab}$
\end{enumerate}
\item (Some important expansions) Expand each of the following with the distributive property.
\begin{enumerate}
\item $(a + b)^2$
\item $(a - b)^2$
\item $(a - b)(a + b)$
\end{enumerate}
\item Factor each of the following as much as possible.
\begin{enumerate}
\item $2x^2 - 18x$
\item $6x^2y + 7xy^2$
\item $-63a^2b - 54ab^2 + 45ab + 36b$
\item $a^2 - b^2$
\item $u^2 - 2uv + v^2$
\item $p^2 + 2pq + q^2$
\end{enumerate}
\item Simplify each of the following expressions as much as possible. For expressions that start off as multiple fractions, you should end up with a single fraction.
\begin{enumerate}
\item $x + 1 + \dfrac{1}{x} - \dfrac{1}{x^2}$
\item $\displaystyle\frac{x}{y} + \frac{y}{x}$
\item $\dfrac{xy^2 - x^2y^3}{3 - 3xy}$
\item $\displaystyle\frac{a^2 - 2ab + b^2}{a^2 - b^2} + \frac{ab + b^2}{a^2 + 2ab + b^2}$
\end{enumerate}
\item Expand $(m + 2n)^2$.
\item (Sophie Germain's identity) Factor $a^4 + 4b^4$.\par\textit{Hint: $a^4 + 4b^4 = (a^4 + 4a^2b^2 + 4b^4) - 4a^2b^2$.}
\end{enumerate}

\newpage
\section{Solving Equations}

\begin{enumerate}
\setcounter{enumi}{9}
\item Solve for the variable in each of the following equations:
\begin{enumerate}
\item $3x + 5 = 11$
\item $6(a - 2) = 3(-a + 4)$
\item $3(2c + 7) - 6(c + 1) = 11$
\item $-4(-t - 2) + t = 5(t + 5) - 17$
\end{enumerate}
\item In each of the following equations, solve for the indicated variable in terms of the others:
\begin{enumerate}
\item $x + 7y = 5$ for $x$
\item $x + 7y = 5$ for $y$
\item $ax^2 + bx + c = 0$ for $b$
\item $pqr + pq + pr + qr = p + q + r + 1$ for $p$
\end{enumerate}
\item Solve each of the following systems of equations for $x$ and $y$:
\begin{enumerate}
\item $x + 2y = 3$ and $3x - y = 2$
\item $-2x + 4y = 7$ and $x - 2y = 3$
\item $4x + y = -2$ and $x + y = 1$
\item $7x - 5y = 1$ and $14x - 10y = 2$
\item $3x + 4y = 25$ and $4x + 3y = 24$
\end{enumerate}
Each of these systems is of the form
\begin{equation*}
Ax + By = E\quad\text{and}\quad Cx + Dy = F.
\end{equation*}
How does the value of $\lvert AD - BC\rvert$ relate to the number of solutions to the system?
\item Eleven years ago Margot was five times as old as Preeti. In sixteen years, Margot will be twice as old as Preeti. How old is Preeti now? \textit{We did this problem in Lesson 3 by using just one variable, but we can also solve it by defining two variables.}
\item Alex buys 3 pencils, 1 binder, and 1 calculator for \$133. Bob buys 1 pencil, 3 binders, and 1 calculator for \$135. Carol buys 1 pencil, 1 binder, and 3 calculators for \$392.
\begin{enumerate}
\item How much would it cost to buy 1 pencil, 1 binder, and 1 calculator?
\item How much does each item cost?
\end{enumerate}
\end{enumerate}

\newpage
\section{Ratios, Proportions, and Percents}

\begin{enumerate}
\setcounter{enumi}{14}
\item Compute each of the following:
\begin{enumerate}
\item $50\%$ of $14$
\item $75\%$ of $24$
\item $20\%$ less than $65$
\item $30\%$ more than $70$
\item $25\%$ of $19\%$ of $400$
\item $8\%$ of $25$
\item $33\%$ more than $300$
\end{enumerate}
\item A bag has two types of food, nuts and dry fruit. The ratio of the weight of nuts in the bag to the weight of dry fruit is $5:4$.
\begin{enumerate}
\item Let $N$ be the weight of nuts in the bag, in grams, and let $F$ be the weight of dry fruit in the bag, also in grams. Write down an equation involving $N$ and $F$.
\item Suppose the total weight of food in the bag is $765$ grams. How many grams of dry fruit are in the bag?
\end{enumerate}
\item A recipe calls for a $7:6:5$ ratio of soy sauce to sugar to sesame oil. If Lena uses 14 teaspoons of sugar, then how much soy sauce and sesame oil should she use (in teaspoons)?
\item (Calculator) The speed of light is approximately $3\times 10^8$ meters per second.
\begin{enumerate}
\item There are roughly $1.6$ kilometers per mile. Find the speed of light in miles per second.
\item The approximation from part (a) is roughly $0.8\%$ too high. Using this, adjust your answer to get a better approximation to the speed of light in miles per second.
\end{enumerate}
\item It takes $3$ identical robots $8$ hours to paint $4$ fences.
\begin{enumerate}
\item How much of a fence can one of these robots paint in one hour?
\item How long, in minutes, would it take $5$ of these robots to paint $7$ fences?
\end{enumerate}
\item Pipe A takes $3$ hours to fill a pool while pipe B takes $4$ hours to fill the pool. Drain C takes $2$ hours to drain the pool. The pool is initially empty, with drain C covered, when pipe A is turned on. One hour later, the cover of drain C is accidentally dislodged, so water begins to drain at the same time. Another hour after that, pipe B is turned on as well. From the point in time that pipe B is turned on, how many hours does it take to fill the pool?
\item On a road trip, Jerry drives for 200 miles at 50 miles per hour, then 108 miles at 36 miles per hour, then some number of miles at 60 miles per hour, and the rest of the trip at 41 miles per hour. The overall distance traversed was 450 miles, and the average speed for the entire trip was 45 miles per hour. How much time did Jerry spend going at 41 miles per hour?
\item (Simpson's paradox) In a certain country, operations are separated into two types, A and B. At Hospital X, type A operations succeed 90\% of the time while type B operations succeed 60\% of the time. At Hospital Y, type A operations succeed 80\% of the time while type B operations succeed 40\% of the time. The overall success rate of operations at Hospital X is 70\% while the overall success rate at Hospital Y is 72\%.
\begin{enumerate}
\item How is it possible that the overall success rate at Hospital Y is higher, even though it is lower for each individual type of operation?
\item If you had to go to one of these hospitals for an operation, which one would you prefer?
\end{enumerate}
\item Katherine takes out a loan of \$100 on January 1, 2000. On January 1 of each subsequent year, the loan balance (i.e. the amount that needs to be paid back) is increased by 20\%.
\begin{enumerate}
\item If Katherine waits until December 1, 2002 to pay back the loan, then how much does she end up paying?
\item Suppose Katherine pays back \$50 on December 1, 2000, another \$50 on December 1, 2001, and the rest on December 1, 2002. How much does she end up paying in total? 
\item (Calculator) Suppose Katherine pays back $\$X$ on December 1 of 2000, 2001, and 2002, and this pays back the loan in full. To the nearest hundredth, what is $X$?
\end{enumerate}
\item (Classic) An orchestra of 120 players takes 75 minutes to play Beethoven's ninth symphony. How long would it take an orchestra of 60 players to play the symphony?
\end{enumerate}

\newpage
\section{Graphing}

\begin{enumerate}
\setcounter{enumi}{24}
\item Draw graphs for each of the following equations:
\begin{enumerate}
\item $2x + 3y = 6$
\item $y = (1/2)x + 3$
\item $x = -y - 1$
\item $y = \lvert x\rvert$
\item $x = y^2 + 2$
\item (Challenge) $y^2 = x(x + 1)(x - 1)$
\end{enumerate}
\item \begin{enumerate}
\item When solving a system of two equations for variables $x$ and $y$, Polly finds that the graphs of the equations are parallel lines. How many solutions are there to the system?
\item For another system, Quaresma finds that the graphs of the equations are the same line. How many solutions are there to this second system?
\end{enumerate}
\item What are the slope and the intercepts of the line given by the equation $4x - 7y = 84$?
\item Let $O = (0,0)$, $P = (14,0)$, and $Q = (5,12)$.
\begin{enumerate}
\item What are the side lengths of triangle $OPQ$?
\item What are the slopes of the sides of triangle $OPQ$?
\end{enumerate}
\item (Centroid) Let $A = (5,1)$, $B = (-1,3)$, and $C = (3,7)$.
\begin{enumerate}
\item Suppose $D$, $E$, and $F$ are the midpoints of $\overline{BC}$, $\overline{CA}$, and $\overline{AB}$, respectively. What are the coordinates of $D$, $E$, and $F$?
\item Write down equations for lines $\overleftrightarrow{AD}$, $\overleftrightarrow{BE}$, and $\overleftrightarrow{CF}$.
\item These three lines intersect at a single point (draw them!), called the \emph{centroid} of triangle $ABC$. What are the coordinates of the centroid?
\end{enumerate}
\item Let $A = (0,2)$ and $B = (12,8)$.
\begin{enumerate}
\item Let $P = (x,y)$. Write down expressions for $(AP)^2$ and $(BP)^2$.
\item If $AP = BP$, then $P$ must lie on a certain line. What are the slope and $y$-intercept of that line?
\item Where does that line intersect $\overline{AB}$?
\end{enumerate}
\end{enumerate}

\newpage
\section{Answers}

\begin{multicols}{2}
\begin{enumerate}
\item \begin{enumerate}
\item $n = 16$
\item $4$
\item $472$
\item greater than 0
\item greater than 1
\item $-2$
\item $4$
\item $\dfrac{b^3c^9}{8a^6}$
\end{enumerate}
\item \begin{enumerate}
\item 10
\item $2\sqrt{2}$
\item $6\sqrt{6}$
\item $17\sqrt{7}$
\item $2\sqrt[3]{18}$
\item $3\sqrt{3}$
\item $14\sqrt{2}$
\item $\dfrac{2\sqrt{3542}}{119}$
\end{enumerate}
\item $x = 1$
\item \begin{enumerate}
\item $6x^2y + 7xy^2$
\item $15a^3b^2 - 5a^3b^3$
\item $-6x + 12x^2 + 18x^3$
\item $-4a - 3b + 2ab$
\end{enumerate}
\item \begin{enumerate}
\item $a^2 + 2ab + b^2$
\item $a^2 - 2ab + b^2$
\item $a^2 - b^2$
\end{enumerate}
\item \begin{enumerate}
\item $2x(x - 9)$
\item $xy(6x + 7y)$
\item $9b(-7a^2 - 6ab + 5a + 4)$
\item $(a - b)(a + b)$
\item $(u - v)^2$ or $(v - u)^2$
\item $(p + q)^2$
\end{enumerate}
\item \begin{enumerate}
\item $\dfrac{x^3 + x^2 + x - 1}{x^2}$
\item $\dfrac{x^2 + y^2}{xy}$
\item $\dfrac{xy^2}{3}$
\item $\dfrac{a}{a + b}$
\end{enumerate}
\item $m^2 + 4mn + 4n^2$
\item $(a^2 + 2b^2 - 2ab)(a^2 + 2b^2 + 2ab)$
\item \begin{enumerate}
\item $x = 2$
\item $a = 8/3$
\item no solution
\item all real numbers
\end{enumerate}
\item \begin{enumerate}
\item $x = 5 - 7y$
\item $y = \dfrac{5 - x}{7}$
\item $b = \dfrac{-ax^2 - c}{x}$
\item $p = \dfrac{-qr + q + r + 1}{qr + q + r - 1}$
\end{enumerate}
\item \begin{enumerate}
\item $(x,y) = (1,1)$
\item no solution
\item $(x,y) = (-1,2)$
\item all pairs $(x,y)$ such that $7x - 5y = 1$
\item $(x,y) = (3,4)$
\end{enumerate}
The system has a unique solution if and only if $AD - BC\neq 0$.
\item 20
\item \begin{enumerate}
\item \$132 (Add up all of the purchases)
\item Pencils are \$0.50, binders are \$1.50, calculators are \$130
\end{enumerate}
\item \begin{enumerate}
\item $7$
\item $18$
\item $52$
\item $91$
\item $19$
\item $2$
\item $399$
\end{enumerate}
\item \begin{enumerate}
\item $N/F = 5/4$
\item $340$
\end{enumerate}
\item $49/3$ tsp soy sauce, $35/3$ tsp sesame oil
\item \begin{enumerate}
\item $187\,500$ miles per second
\item $\approx 186\,012$ miles per second
\end{enumerate}
\item \begin{enumerate}
\item $1/6$
\item $504$
\end{enumerate}
\item After one hour, $1/3$ of the pool is full. For the second hour, the pool fills at a rate of $(1/3) - (1/2) = -1/6$ of the pool per hour, so after the second hour, $1/6$ of the pool is full. Afterwards, the fill rate is $(1/3) + (1/4) - (1/2) = 1/12$ of the pool per hour, and we have $5/6$ of the pool left, so the answer is $(5/6) / (1/12) = \boxed{10}$ hours.
\item Let $F$ be the time spent at $60$ miles per hour and $S$ be the time spent at $41$ miles per hour. The total time for the trip was $450 / 45 = 10$ hours, so $F + S = 3$. Also, we need $60F + 41S = 142$. Solving yields $F = 1$ and $S = \boxed{2}$ hours.
\item \begin{enumerate}
\item This can happen when a much higher proportion of Hospital Y's operations are type A (lower risk) compared to Hospital X. Given these numbers, only 50\% of Hospital X's operations are type A, vs 80\% of Hospital Y's.
\item Hospital X
\end{enumerate}
\item \begin{enumerate}
\item $\$144$
\item $\$112$
\item On January 1, 2001, the balance is $1.2(100 - X) = 120 - 1.2X$. Then, on January 1, 2002, the balance is $1.2(120 - 1.2X - X) = 144 - 2.64X$. This is paid off in full by the last $X$, so $144 - 2.64X = X$, so $X\approx\boxed{39.56}$.
\end{enumerate}
\item $75$
\item See \href{https://www.desmos.com/}{Desmos}
\item \begin{enumerate}
\item none
\item infinitely many
\end{enumerate}
\item Slope $4/7$\par $x$-intercept $(21,0)$\par $y$-intercept $(0,-12)$
\item \begin{enumerate}
\item $OP = 14$, $PQ = 15$, $QO = 13$
\item $\overline{OP}$ has slope 0\par $\overline{PQ}$ has slope $-4/3$\par $\overline{QO}$ has slope $12/5$
\end{enumerate}
\item \begin{enumerate}
\item $D = (1,5)$\par $E = (4,4)$\par $F = (2,2)$
\item $\overleftrightarrow{AD}$ is $y = -x + 6$\par $\overleftrightarrow{BE}$ is $y - 3 = (1/5)(x + 1)$\par $\overleftrightarrow{CF}$ is $y = 5x - 8$
\item $(7/3, 11/3)$
\end{enumerate}
\item \begin{enumerate}
\item $(AP)^2 = x^2 + (y - 2)^2$\par $(BP)^2 = (x - 12)^2 + (y - 8)^2$
\item Expand $(AP)^2 = (BP)^2$ and cancel squares: $-4y + 4 = -24x - 16y + 208$. This means $12y = -24x + 204$, so $y = -2x + 17$. The slope is $\boxed{-2}$ and the $y$-intercept is $\boxed{(0,17)}$.
\item $(6,5)$  (This is the midpoint of $\overline{AB}$.)
\end{enumerate}
\end{enumerate}
\end{multicols}


\end{document}
