\section{Complex Numbers II: Exponentials}

For real numbers $\theta$, we define $e^{i\theta} = \cos\theta + i\sin\theta$, with this expression also sometimes denoted $\cis\theta$. For complex numbers $z = x + yi$, we define $e^z = e^x\cdot e^{iy}$, where $e^x$ is the real exponential function with base $e\approx 2.718$ evaluated at $x$.

\subsection{Review problems}

\begin{enumerate}
\item \emph{Magnitude and argument.} For each of the following complex numbers $z$, find the magnitude $\lvert z\rvert$ and all possible values of the argument $\arg z$ (if defined).
\begin{enumerate}
\item 1
\item $2 + 2i$
\item $-3i$
\item $-2 + 2\sqrt{3}i$
\item 0
\end{enumerate}
\item \emph{More on arguments.} For any non-zero real number $z$, we define the \textbf{principal value} of the argument to be the value of $\arg z$ in the interval\footnote{Some authors may opt for another interval like $[0,2\pi)$, but $(-\pi,\pi]$ is the most common.} $(-\pi,\pi]$ and denote this by $\Arg z$.
\begin{enumerate}
\item Let $z$ and $w$ be complex numbers with arguments $2\pi/3$ and $3\pi/4$, respectively. What is (a possible value of) the argument of $z^2/w$?
\item In general, $\arg(zw) = \arg z + \arg w$ is true when interpreted to mean that if $\theta$ is a possible value for the argument of $z$ and $\phi$ is a possible value for the argument of $w$, then $\theta + \phi$ is a possible value for the argument of $zw$.\par
Does $\Arg(zw) = \Arg z + \Arg w$ hold?
\end{enumerate}
\item \emph{Exponential form.} Fill in the table below.
\begin{center}
\begingroup
\setlength{\tabcolsep}{10pt}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|} \hline
Standard form & Exponential form \\ \hline
$7i$ & $7e^{i\pi/2}$ \\ \hline
{} & $e^{\pi i}$ \\ \hline
$1 + i$ & {} \\ \hline
$3 - 3\sqrt{3}i$ & {} \\ \hline
{} & $2e^{-2i\pi/3}$ \\ \hline
{} & $4e^{5i\pi/12}$ \\ \hline
\end{tabular}
\endgroup
\end{center}
\item \emph{Exact values of roots of unity.} For any positive integer $n$, an $\boldsymbol n$\textbf{-th root of unity} is a solution to the equation $z^n = 1$. Compute the $n$-th roots of unity for $n = 1, 2, 3, 4, 6, 8$.
\item \emph{Roots of other complex numbers.} Let $n$ be a positive integer and let $z$ be any complex number. Writing $z$ in exponential form, let real numbers $r\geq 0$ and $\theta$ be such that $z = re^{i\theta}$.
\begin{enumerate}
\item In terms of $r$, $\theta$, and $n$, write down an $n$-th root of $z$.
\item Show that if $w$ is an $n$-th root of $z$ and $\zeta$ is an $n$-th root of unity, then $\zeta w$ is also an $n$-th root of $z$.
\item Conversely, show that if $w_1$ and $w_2$ are two $n$-th roots of $z$, then there is an $n$-th root of unity $\zeta$ for which $w_2 = \zeta w_1$.
\item Let $\zeta = e^{2\pi i/7}$. Express all solutions to $z^7 = 128$ in terms of $\zeta$.
\end{enumerate}
\item \emph{Primitive $n$-th roots of unity.} Let $n$ be a positive integer. A \textbf{primitive $\boldsymbol n$-th root of unity} is an $n$-th root of unity which is not a $k$-th root of unity for any positive integer $k < n$.
\begin{enumerate}
\item For $n = 1, 2, 3, 4, 6, 8$, what are the primitive $n$-th roots of unity?
\item Any $n$-th root of unity can be written in the form $e^{2\pi ik/n}$ for an integer $k$. Show that this is a primitive $n$-th root of unity if and only if $\gcd(k,n) = 1$.
\item Let $\zeta$ be a primitive $n$-th root of unity. Show that for any integer $k$,
\begin{equation*}
1 + \zeta^k + \zeta^{2k} + \cdots + \zeta^{(n - 1)k} = \begin{cases} n & n\mid k, \\ 0 & \text{otherwise}. \end{cases}
\end{equation*}
\item \textit{Roots of unity filter.} Let
\begin{equation*}
f(z) = a_0 + a_1z + a_2z^2 + \cdots + a_{mn}z^{mn}
\end{equation*}
be a polynomial and let $\zeta = e^{2\pi i/n}$. Show that
\begin{equation*}
a_0 + a_n + a_{2n} + \cdots + a_{mn} = \frac{f(1) + f(\zeta) + f(\zeta^2) + \cdots + f(\zeta^{n - 1})}{n}.
\end{equation*}
\item Compute $\binom{2025}{0} + \binom{2025}{3} + \binom{2025}{6} + \cdots + \binom{2025}{2025}$.
\end{enumerate}
\item \emph{Sine and cosine as exponentials.} 
\begin{enumerate}
\item Let $\theta$ be a real number. Show that
\begin{equation*}
\cos\theta = \frac{e^{i\theta} + e^{-i\theta}}{2}\quad\text{and}\quad\sin\theta = \frac{e^{i\theta} - e^{-i\theta}}{2i}.
\end{equation*}
\item For any complex number $z$, we define
\begin{equation*}
\cos z = \frac{e^{iz} + e^{-iz}}{2}\quad\text{and}\quad\sin z = \frac{e^{iz} - e^{-iz}}{2i}.
\end{equation*}
Other trig functions of $z$ are defined in terms of sine and cosine as usual, e.g. $\tan z = \frac{\sin z}{\cos z}$.\par
Calculate $\cos(i)$ and $\sin(i)$.
\item Prove that $\cos^2 z + \sin^2 z = 1$ for all complex numbers $z$.
\item Prove that $2\sin z\cos z = \sin(2z)$ for all complex numbers $z$.
\end{enumerate}
\end{enumerate}


\subsection{Challenge problems}

\begin{enumerate}\setcounter{enumi}{7}
\item Let $\ln:(0,\infty)\to\mathbb{R}$ (temporarily) denote the natural logarithm function, i.e. the inverse of the real exponential function $x\mapsto e^x$. Define $L:\mathbb{C}\backslash\{0\}\to\mathbb{C}$ by
\begin{equation*}
L(z) = \ln\lvert z\rvert + i\arg z.
\end{equation*}
Since $\arg$ is multi-valued, $L$ is also multi-valued, with its possible output values differing by integer multiples of $2\pi i$.
\begin{enumerate}
\item Show that $e^{L(z)} = z$ (no matter what value of $\arg z$ we take). As such, $L$ acts as an ``inverse'' to the complex exponential, so we call it the \textbf{complex logarithm} and denote it by $\log z$ instead of $L(z)$. We define the \textbf{principal value of the logarithm} to be
\begin{equation*}
\Log z = \ln\lvert z\rvert + i\Arg z.
\end{equation*}
\item Compute the following:
\begin{enumerate}
\item $\Log 1$
\item $\Log i$
\item $\Log (-3 - 3i)$
\end{enumerate}
\item Prove that $\log(zw) = \log z + \log w$ for all $z,w\in\mathbb{C}\backslash\{0\}$.\par
Is it true that $\Log(zw) = \Log z + \Log w$ for all $z,w\in\mathbb{C}\backslash\{0\}$?
\item Show that the solutions to $\tan w = z$ are
\begin{equation*}
w = \frac{1}{2i}\log\left(\frac{1 + iz}{1 - iz}\right).
\end{equation*}
This makes the expression on the right hand side the (multi-valued) \textbf{inverse tangent function} for complex numbers.
\end{enumerate}
\item The \textbf{fundamental theorem of algebra} states that for every non-constant one-variable polynomial function $p(z) = a_0 + a_1z + \cdots + a_nz^n$, where $n\geq 1$ and the $a_i$ are complex numbers with $a_n\neq 0$, there exists $r\in\mathbb{C}$ with $p(r) = 0$. In this problem, we provide one proof of this theorem. Note that without loss of generality, we can suppose $a_n = 1$, and since $p(0) = 0$ when $a_0 = 0$, we only need to address the case that $a_0\neq 0$.
\begin{enumerate}
\item Let $R = 2(\lvert a_0\rvert + \lvert a_1\rvert + \cdots + \lvert a_{n - 1}\rvert + 1)$. Show that if $\lvert z\rvert\geq R$, then 
\begin{equation*}
\left\lvert\frac{p(z)}{z^n}\right\rvert\geq\frac{1}{2}.
\end{equation*}
\item A theorem of mathematical analysis tells us that since $\lvert p(z)\rvert$ is a continuous function of $z$, there is a point $z_0$ with $\lvert z_0\rvert\leq R$ for which $m = \lvert p(z_0)\rvert$ is the minimum value of $\lvert p(z)\rvert$ as $z$ ranges over all complex numbers with magnitude at most $R$. Show that in fact, $m$ is the minimum value of $\lvert p(z)\rvert$ as $z$ ranges over all complex numbers.
\item Suppose for the sake of contradiction that $m > 0$. By translating, we can suppose that $z_0 = 0$, so that $m = \lvert a_0\rvert$. Let $0 < k\leq n$ be the smallest positive integer with $a_k\neq 0$, let $\omega$ be any complex number satisfying $\omega^k = -a_0/a_k$, and let 
\begin{equation*}
\epsilon = \frac{1}{2}\min\left(1, \lvert a_0\rvert\cdot\left[\lvert a_{k + 1}\omega^{k + 1}\rvert + \lvert a_{k + 2}\omega^{k + 2}\rvert + \cdots + \lvert a_n\omega^n\rvert\right]^{-1}\right).
\end{equation*}
Show that
\begin{equation*}
\lvert p(\epsilon\cdot\omega) - (a_0 - a_0\epsilon^k)\rvert\leq\frac{\lvert a_0\rvert}{2}\epsilon^k,
\end{equation*}
and hence deduce that $\lvert p(\epsilon\cdot\omega)\rvert < m$, contradicting minimality of $m$.
\end{enumerate}
\item Let $\zeta_1, \ldots, \zeta_{\varphi(n)}$ be all the primitive $n$-th roots of unity (where $\varphi(n)$ is the number of positive integers $k\leq n$ satisfying $\gcd(k,n) = 1$). We define the $\boldsymbol n$\textbf{-th cyclotomic polynomial} to be
\begin{equation*}
\Phi_n(X) = (X - \zeta_1)(X - \zeta_2)\cdots (X - \zeta_{\varphi(n)}).
\end{equation*}
\begin{enumerate}
\item Compute $\Phi_n(X)$ for $n = 1, 2, 3, 4, 6, 8$.
\item Show that $X^n - 1 = \prod_{d\mid n}\Phi_d(X)$.
\item Show that $\Phi_p(X) = X^{p - 1} + X^{p - 2} + \cdots + X + 1$ whenever $p$ is prime.
\item An important fact for number theory is that the cyclotomic polynomials are \textbf{irreducible} (over the integers), meaning that they cannot be written as products of polynomials of lower degree with integer coefficients. We will not go through a general proof here, but some special cases are easier to tackle.\par
Let $p$ be a prime. By considering a shifted cyclotomic polynomial $f(X) = \Phi_p(X + 1)$, or otherwise, show that $\Phi_p(X)$ is irreducible.
\end{enumerate}
\end{enumerate}


\newpage
\subsection{Answers}

\begin{enumerate}
\item In all of the below, $k$ can be any integer.
\begin{enumerate}
\item $\lvert 1\rvert = 1$ and $\arg 1 = 2\pi ik$
\item $\lvert 2 + 2i\rvert = 2\sqrt{2}$ and $\arg(2 + 2i) = \pi/4 + 2\pi ik$
\item $\lvert -3i\rvert = 3$ and $\arg(-3i) = -\pi/2 + 2\pi ik$
\item $\lvert -2 + 2\sqrt{3}i\rvert = 4$ and $\arg(-2 + 2\sqrt{3}i) = 2\pi/3 + 2\pi ik$
\item $\lvert 0\rvert = 0$ and $\arg 0$ is undefined
\end{enumerate}
\item \begin{enumerate}
\item $2\cdot\frac{2\pi}{3} - \frac{3\pi}{4} = \frac{7\pi}{12}$ (any integer multiple of $2\pi$ can be added to this)
\item Let $z = w = -i$. Then $\Arg z = \Arg w = -\pi/2$, but
\begin{equation*}
\Arg(zw) = \Arg(-1) = \pi\neq\Arg z + \Arg w.
\end{equation*}
\end{enumerate}
\item When converting from standard form to exponential form, we can add any integer multiple of $2\pi i$ to the exponent to get another valid exponential form expression.
\begin{center}
\begingroup
\setlength{\tabcolsep}{10pt}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|} \hline
Standard form & Exponential form \\ \hline
$7i$ & $7e^{i\pi/2}$ \\ \hline
\color{red} $-1$ & $e^{\pi i}$ \\ \hline
$1 + i$ & \color{red} $\sqrt{2}e^{i\pi/4}$ \\ \hline
$3 - 3\sqrt{3}i$ & \color{red} $6e^{-i\pi/3}$ \\ \hline
\color{red} $-1 - \sqrt{3}i$ & $2e^{-2i\pi/3}$ \\ \hline
\color{red} $(\sqrt{6} - \sqrt{2}) + (\sqrt{6} + \sqrt{2})i$ & $4e^{5i\pi/12}$ \\ \hline
\end{tabular}
\endgroup
\end{center}
\item See the table below.
\begin{center}
\begingroup
\setlength{\tabcolsep}{10pt}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|} \hline
$n$ & $n$-th roots of unity \\ \hline
1 & 1 \\ \hline
2 & 1, $-1$ \\ \hline
3 & 1, $\frac{-1 + \sqrt{3}i}{2}$, $\frac{-1 - \sqrt{3}i}{2}$ \\ \hline
4 & 1, $i$, $-1$, $-i$ \\ \hline
6 & 1, $\frac{1 + \sqrt{3}i}{2}$, $\frac{-1 + \sqrt{3}i}{2}$, $-1$, $\frac{-1 - \sqrt{3}i}{2}$, $\frac{1 - \sqrt{3}i}{2}$ \\ \hline
8 & 1, $\frac{1 + i}{\sqrt{2}}$, $i$, $\frac{1 - i}{\sqrt{2}}$, $-1$, $\frac{-1 - i}{\sqrt{2}}$, $-i$, $\frac{1 - i}{\sqrt{2}}$ \\ \hline
\end{tabular}
\endgroup
\end{center}
\item \begin{enumerate}
\item $r^{1/n}e^{i(\theta/n)}$
\item If $w^n = z$ and $\zeta^n = 1$, then $(\zeta w)^n = \zeta^nw^n = 1\cdot z = z$.
\item If $w_1^n = w_2^n = z$, then $(w_1/w_2)^n = 1$, so $w_1/w_2 = \zeta$ for some $n$-th root of unity $\zeta$.
\item $z = 2\zeta^k$ for $k = 0, 1, \ldots, 6$.
\end{enumerate}
\item \begin{enumerate}
\item See the table below.
\begin{center}
\begingroup
\setlength{\tabcolsep}{10pt}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|} \hline
$n$ & primitive $n$-th roots of unity \\ \hline
1 & 1 \\ \hline
2 & $-1$ \\ \hline
3 & $\frac{-1 + \sqrt{3}i}{2}$, $\frac{-1 - \sqrt{3}i}{2}$ \\ \hline
4 & $i$, $-i$ \\ \hline
6 & $\frac{1 + \sqrt{3}i}{2}$, $\frac{1 - \sqrt{3}i}{2}$ \\ \hline
8 & $\frac{1 + i}{\sqrt{2}}$, $\frac{1 - i}{\sqrt{2}}$, $\frac{-1 - i}{\sqrt{2}}$, $\frac{1 - i}{\sqrt{2}}$ \\ \hline
\end{tabular}
\endgroup
\end{center}
\item Suppose that $\gcd(k,n) = 1$ and $(e^{2\pi ik/n})^m = e^{2\pi ikm/n} = 1$ for some positive integer $m$. Then $km/n$ must be an integer, and with $\gcd(k,n) = 1$, this forces $m$ to be divisible by $n$. Hence $e^{2\pi ik/n}$ is not an $m$-th root of unity for any $m < n$.\par
Inversely, suppose $\gcd(k,n) = d > 1$. Then $(e^{2\pi ik/n})^{n/d} = e^{2\pi i(k/d)} = 1$, so $e^{2\pi ik/n}$ is an $(n/d)$-th root of unity and hence not a primitive $n$-th root of unity.
\item If $n$ divides $k$, then $\zeta^k = 1$ so the sum evaluates to $n$. If $n$ does not divide $k$, then since $\zeta$ is a primitive $n$-th root of unity, $\zeta^k\neq 1$. Therefore,
\begin{equation*}
1 + \zeta^k + \cdots + \zeta^{(n - 1)k} = \frac{1 - \zeta^{nk}}{1 - \zeta^k} = \frac{1 - 1}{1 - \zeta^k} = 0.
\end{equation*}
\item We start by writing the right hand side as
\begin{equation*}
\frac{1}{n}\sum_{k = 0}^{n - 1}f(\zeta^k) = \frac{1}{n}\sum_{k = 0}^{n - 1}\sum_{j = 0}^{mn} a_j(\zeta^k)^j = \frac{1}{n}\sum_{j = 0}^{mn}\left(a_j\sum_{k = 0}^{n - 1}(\zeta^k)^j\right).
\end{equation*}
The inner sum, by part (c), is $n$ whenever $n\mid j$ and $0$ otherwise, so the sum becomes
\begin{equation*}
\frac{1}{n}(na_0 + na_n + na_{2n} + \cdots + na_{mn}) = a_0 + a_n + \cdots + a_{mn}.
\end{equation*}
\item Let $\zeta = e^{2\pi i/3}$ and let $f(z) = (1 + z)^{2025}$, so that $a_j = \binom{2025}{j}$. Then
\begin{equation*}
\binom{2025}{0} + \binom{2025}{3} + \cdots + \binom{2025}{2025} = \frac{f(1) + f(\zeta) + f(\zeta^2)}{3} = \frac{2^{2025} - 2}{3}.
\end{equation*}
\end{enumerate}
\item \begin{enumerate}
\item We have $e^{i\theta} = \cos\theta + i\sin\theta$ and
\begin{equation*}
e^{-i\theta} = \cos(-\theta) + i\sin(-\theta) = \cos\theta - i\sin\theta.
\end{equation*}
Adding the equations and dividing by 2 gives us the cosine formula, while subtracting the equations and dividing by $2i$ gives us the sine formula.
\item $\cos i = \frac{e + e^{-1}}{2}$ and $\sin i = \frac{e^{-1} - e}{2i} = \frac{(e - e^{-1})i}{2}$
\item $\cos^2 z + \sin^2 z = \left(\frac{e^{iz} + e^{-iz}}{2}\right)^2 + \left(\frac{e^{iz} - e^{-iz}}{2i}\right)^2 = \frac{e^{2iz} + 2 + e^{-2iz}}{4} - \frac{e^{2iz} - 2 + e^{-2iz}}{4} = 1$
\item $2\sin z\cos z = 2\left(\frac{e^{iz} - e^{-iz}}{2i}\right)\left(\frac{e^{iz} + e^{-iz}}{2}\right) = \frac{e^{2iz} - e^{-2iz}}{2i} = \sin(2z)$
\end{enumerate}
\item \begin{enumerate}
\item $e^{L(z)} = e^{\ln\lvert z\rvert}e^{i\arg z} = \lvert z\rvert e^{i\arg z} = z$
\item \begin{enumerate}
\item 0
\item $\pi i/2$
\item $\ln(3\sqrt{2}) - 3\pi i/4$
\end{enumerate}
\item $\log(zw) = \ln\lvert zw\rvert + i\arg(zw) = (\ln\lvert z\rvert + \ln\lvert w\rvert) + i(\arg z + \arg w) = \log z + \log w$\par
The statement $\Log(zw)\stackrel{?!}{=} \Log z + \Log w$ fails for the same reason that $\Arg(zw)\stackrel{?!}{=}\Arg z + \Arg w$ fails. In particular, when $\lvert z\rvert = 1$, we have $\Log z = i\Arg z$, so we can just take $z = w = -i$ as in our counterexample for problem 2(b).
\item We have
\begin{equation*}
\tan w = \frac{\sin w}{\cos w} = \frac{1}{i}\frac{e^{iw} - e^{-iw}}{e^{iw} + e^{-iw}} = \frac{1}{i}\frac{e^{2iw} - 1}{e^{2iw} + 1}.
\end{equation*}
If $\tan w = z$, then
\begin{align*}
\frac{1}{i}\frac{e^{2iw} - 1}{e^{2iw} + 1} &= z, \\
e^{2iw} - 1 &= iz(e^{2iw} + 1), \\
(1 - iz)e^{2iw} &= 1 + iz, \\
e^{2iw} &= \frac{1 + iz}{1 - iz}, \\
w &= \frac{1}{2i}\log\left(\frac{1 + iz}{1 - iz}\right).
\end{align*}
\end{enumerate}
\item \begin{enumerate}
\item By the triangle inequality,
\begin{equation*}
\left\lvert\frac{p(z)}{z^n}\right\rvert = \left\lvert 1 + \sum_{j = 0}^{n - 1}\frac{a_j}{z^{n - j}}\right\rvert\geq 1 - \left\lvert\sum_{j = 0}^{n - 1}\frac{a_j}{z^{n - j}}\right\rvert,
\end{equation*}
so it suffices to show the last term is at most $1/2$. Since $\lvert z\rvert\geq 1$,
\begin{equation*}
\left\lvert\sum_{j = 0}^{n - 1}\frac{a_j}{z^{n - j}}\right\rvert\leq\sum_{j = 0}^{n - 1}\frac{\lvert a_j\rvert}{\lvert z\rvert^{n - j}}\leq\frac{1}{\lvert z\rvert}\sum_{j = 0}^{n - 1}\lvert a_j\rvert.
\end{equation*}
That this is at most $1/2$ then follows from $\lvert z\rvert\geq R > 2\sum_{j = 0}^{n - 1}\lvert a_j\rvert$.
\item We have $m\leq\lvert p(z)\rvert$ whenever $\lvert z\rvert\leq R$, so in particular $m\leq\lvert p(0)\rvert = \lvert a_0\rvert$. When $\lvert z\rvert\geq R$, part (a) gives us $\lvert p(z)\rvert\geq R^n/2$. If $m\leq 1$, then $R^n/2\geq 2^{n - 1}\geq m$, and if $m > 1$, then $R^n/2\geq\lvert a_0\rvert^n > \lvert a_0\rvert\geq m$ as well. Hence $m$ is indeed the minimum of $\lvert p(z)\rvert$ over all $z$.
\item We start by writing
\begin{equation*}
p(\epsilon\cdot\omega) = a_0 + a_k\epsilon^k\omega^k + \sum_{j = k + 1}^n a_j\epsilon^j\omega^j = a_0 - a_0\epsilon^k + \epsilon^k\sum_{j = k + 1}^n a_j\epsilon^{j - k}\omega^j.
\end{equation*}
Since $\epsilon\leq 1/2 < 1$, the sum can be bounded by
\begin{equation*}
\left\lvert\sum_{j = k + 1}^n a_j\epsilon^{j - k}\omega^j\right\rvert\leq\sum_{j = k + 1}^n \lvert a_j\omega^j\rvert\cdot\epsilon^{j - k}\leq\epsilon\sum_{j = k + 1}^n\lvert a_j\omega^j\rvert.
\end{equation*}
This last expression is bounded above by $\lvert a_0\rvert/2$, as required. To finish,
\begin{equation*}
\lvert p(\epsilon\cdot\omega)\rvert\leq\lvert a_0\rvert - \lvert a_0\rvert\epsilon^k + \frac{\lvert a_0\rvert}{2}\epsilon^k = m\left(1 - \frac{\epsilon^k}{2}\right) < m.
\end{equation*}
\end{enumerate}
\item \begin{enumerate}
\item See the table below.
\begin{center}
\begingroup
\setlength{\tabcolsep}{10pt}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|} \hline
$n$ & $\Phi_n(X)$ \\ \hline
1 & $X - 1$ \\ \hline
2 & $X + 1$ \\ \hline
3 & $X^2 + X + 1$ \\ \hline
4 & $X^2 + 1$ \\ \hline
6 & $X^2 - X + 1$ \\ \hline
8 & $X^4 + 1$ \\ \hline
\end{tabular}
\endgroup
\end{center}
\item First, note that when $d\neq d'$, the polynomials $\Phi_d(X)$ and $\Phi_{d'}(X)$ share no roots because a primitive $d$-th root of unity cannot also be a $(d')$-th root of unity or vice versa. Therefore, all roots of the right hand side occur with multiplicity 1, as is the case on the left hand side. As such, it suffices to show that the left and right hand sides have the same roots.\par
If $\zeta$ is a root of the right hand side, then $\zeta$ is a root of $\Phi_d(X)$ for some $d\mid n$. Since $\zeta^d = 1$ and $d\mid n$, we also have $\zeta^n = 1$, so $\zeta$ is a root of the left hand side.\par
Conversely, let $\zeta$ be a root of $X^n - 1$ and suppose $\zeta$ is a primitive $d$-th root of unity. Let $n = qd + r$, where $0\leq r < d$ is the remainder when $n$ is divided by $d$. Then $\zeta^(n - qd) = \zeta^r = 1$. Since $\zeta$ is a primitive $d$-th root of unity, we must have $r = 0$, so $d\mid n$. This means $\zeta$ is a root of $\Phi_d(X)$ and hence a root of the right hand side.
\item When $p$ is prime, $X^p - 1 = \Phi_1(X)\Phi_p(X)$, so
\begin{equation*}
\Phi_p(X) = \frac{X^p - 1}{X - 1} = X^{p - 1} + \cdots + X + 1.
\end{equation*}
\item For $p = 2$, we saw that $\Phi_2(X) = X + 1$, which cannot be factored, so for the rest of the argument, we can suppose $p > 2$. Any factorisation of $f(X) = \Phi_p(X + 1)$ yields a factorisation of $\Phi_p(X)$, so it suffices to show that $f(X)$ is irreducible. We calculate
\begin{equation*}
f(X) = \frac{(X + 1)^p - 1}{(X + 1) - 1} = X^{p - 1} + pX^{p - 2} + \binom{p}{2}X^{p - 3} + \cdots + \binom{p}{p - 2}X + p.
\end{equation*}
Suppose for the sake of contradiction that we have a (non-trivial) factorisation
\begin{equation*}
f(X) = (X^k + a_{k - 1}X^{k - 1} + \cdots + a_0)(X^{p - 1 - k} + b_{p - 2 - k}X^{p - 2 - k} + \cdots + b_0),
\end{equation*}
where $0 < k < p - 1$. Expanding, $a_0b_0 = p$, so we can suppose without loss of generality that $a_0 = \pm p$ and $b_0 = \pm 1$; whatever the signs are, $p\mid a_0$ but $p\nmid b_0$. Looking at the linear term, $a_0b_1 + a_1b_0 = \binom{p}{p - 2}$, which is divisible by $p$. Since $p\mid a_0$ but $p\nmid b_0$, we must have $p\mid a_1$. Looking at the quadratic term, $a_0b_2 + a_1b_1 + a_2b_0 = \binom{p}{p - 3}$, which is divisible by $p$. Since $p\mid a_0, a_1$ but $p\nmid b_0$, we must have $p\mid a_2$. Continuing in this way, $p\mid a_j$ for all $0\leq j\leq k - 1$. (If $k\geq p - 1 - k$, then we take $b_{p - 1 - k} = 1$ and $b_j = 0$ for all $j\geq p - k$.) The coefficient of $X^k$ is then $a_0b_k + a_1b_{k - 1} + \cdots + a_{k - 1}b_1 + b_0$, which is not divisible by $p$. However, the coefficient of $X^k$ should also be $\binom{p}{p - 1 - k}$, which is divisible by $p$ since $0 < k < p - 1$. This is a contradiction, so $f(X)$ cannot factor non-trivially over the integers.
\end{enumerate}
\end{enumerate}